---
title: "Microarray Data - H.pylori infection and IFN treatment - DGE Reanalysis with LIMMA"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, echo=FALSE, results='hide', messages=FALSE}
rm(list=ls())
suppressMessages(library(limma))
suppressMessages(library(xlsx))
suppressMessages(library(pheatmap))
library(xtable)
library(reshape2)

root_folder = normalizePath("../../")

result_folder = file.path(root_folder, "Results/Microarray")
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

load(file.path(root_folder, "Data/Processed/Microarray/Micro_array_preprocessed_data.Rdata"))
```

# Introduction

This is data from a mouse stomach samples of mouse experiments (H.pylori infection, IFNy treatment) hybridized as dual channel libraries to Agilent 014868 Whole genome mouse 4x44 micro arrays. 


# DGE

We here use standard two-channel analysis of the microarray data and compare:
  - gastric epithelial cells from mice infected with H.pylori and controls
  - IFN treated vs. control epithelial organoids 

```{r}
all_results = list()
```


```{r}
controls = MA$genes$ControlType!=0

MA.avg <- avereps(MA[!controls,], ID=MA[!controls,]$genes$ProbeName)

dswp = ed$dye_swap
cy3_label = ed$Cy3
cy5_label = ed$Cy5
scan_id = ed$Chip.Barcode
ed$array_label = paste(ifelse(dswp, cy3_label, cy5_label), "_",ifelse(dswp,cy5_label, cy3_label),"_",dswp,"_",ed$Scan.ID,sep="")

sel_samples = rownames(subset(ed, Experiment=="IFN"))
ed_tmp = ed[sel_samples,]
design = ifelse(ed_tmp$Cy5_treatment=="IFN", 1, -1)
fit <- lmFit(MA.avg[,sel_samples], design)
fit <- eBayes(fit)
res = topTable(fit, adjust = "BH",number=nrow(fit), confint = T)
ratio_tab = sweep(MA.avg[,sel_samples]$M, 2, design, "*")
colnames(ratio_tab) = ed_tmp$array_label
res = merge(res, ratio_tab, all.x=T, by.x="ProbeName", by.y=0, sort=F)
all_results[["IFN_vs_ctrl_organoids"]] = res

sel_samples = rownames(subset(ed, Experiment=="Infection Antrum"))
ed_tmp = ed[sel_samples,]
design = ifelse(ed_tmp$Cy5_infection=="Hpy", 1, -1)
fit <- lmFit(MA.avg[,sel_samples], design)
fit <- eBayes(fit)
res = topTable(fit, adjust = "BH",number=nrow(fit), confint = T)
ratio_tab = sweep(MA.avg[,sel_samples]$M, 2, design, "*")
colnames(ratio_tab) = ed_tmp$array_label
res = merge(res, ratio_tab, all.x=T, by.x="ProbeName", by.y=0, sort=F)
all_results[["Hpy_vs_NI"]] = res

common_colnames = colnames(topTable(fit, adjust = "BH",number=nrow(fit)))
```

## Volcano plot

```{r, volcano, echo=FALSE, fig.width=8, fig.height=6}
all_target_conditions = names(all_results)
par(mfrow=c(1,2), mar=c(4,4,4,2))

for (tc in all_target_conditions) {
  r = all_results[[tc]]
  plot(r$logFC, -log10(r$adj.P.Val),xlab="log2 Fold Change",ylab="-log10(adj. p-val)", ylim=c(0,max(2,max(-log10(r$adj.P.Val),na.rm=T))))
  title(main=tc, sub=paste("(",nrow(subset(r, adj.P.Val < 0.05))," signif. DE genes)",sep="") )
  abline(h=-log10(0.05),col="red")
  abline(v=c(-1,1))
}
```

# Data export 

All result files are exported to `r result_folder`.


```{r, write_tabs, echo=FALSE}
output_file_prefix = paste(result_folder,"Differential_expression_results_", sep="/")
selected_cols = c("ProbeName", "GeneSymbol", "GeneName","EntrezID","GeneSymbol_Gencode","GencodeM12MappingStatus","Transcript_Type","logFC","AveExpr","t","P.Value","adj.P.Val" )

for (tc in all_target_conditions) {
  #write.table(all_results[[tc]][,selected_cols], file=paste(output_file_prefix, tc, ".txt", sep="" ), row.names=F , sep="\t", dec=".")
  write.table(all_results[[tc]], file=paste(output_file_prefix, tc, ".txt", sep="" ), row.names=F , sep="\t", dec=".")
}

```


```{r}
filename = paste(result_folder,"DGE_results.Rdata",sep="/")
exp_design = ed
save(all_results, exp_design, MA,file=filename)
```


# Session Info
```{r}
sessionInfo()
```